---
title: "EDA of Mortgage Loan Decisions"
author: "Afsar Ali"
output:
  html_document:
    toc: yes
    toc_depth: 4
---

## Code header 

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# Course: ECON 5300L Applied Econometric
# Title: EDA of Mortgage Loan Decisions 
# Purpose: Explore Mortgage Loan Decisions outcome
# Data: MLD Data file.csv
# Date: Mar 26, 2018
# Author: Afsar Ali
```

#Exploratory Data Analysis (EDA)

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
rm(list=ls(all=TRUE))  # Clear all data in environment

#load Packages 
library(tidyverse)
library(GGally)
library(plotly)
library(car)
library(ggplot2)
library(stargazer)
library(sandwich)
library(lmtest)   # waldtest; see also coeftest.
library(psych)
library(aod)
library(Rcpp)
library(Hmisc)
library(pastecs)
library(popbio)
library(rms)
library(Hmisc)
library(kableExtra)

# Load data 
datd <- read.csv("MLD Data File.csv", header = TRUE)
#Review selected data
glimpse(datd)
kable(summary(datd))
```

## Data Discription

In light of the relatively small number of mortgage loan applications made by minorities, these extra variables were collected for all applications by blacks and Hispanics and for a random sample. The data set includes the following variables.

- APPROVE = 1 if mortgage loan was approved, = 0 otherwise
- GDLIN = 1 if credit history meets guidelines, = 0 otherwise
- LOANPRC = loan amount/purchase price
- OBRAT = other obligations as a percent of total income
- MALE = 1 if male, = 0 otherwise
- MARRIED = 1 if married, = 0 otherwise
- BLACK = 1 if black, = 0 otherwise
- HISPAN = 1 if Hispanic, = 0 otherwise

##Remove 20 na's obs based on above summary 

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#filter data 
dats <-
  datd %>%
  filter(GDLIN != 666) %>%
  filter(MALE != ".") %>%
  filter(MARRIED != ".") %>%
  mutate(LOANPRC = LOANPRC*100)

# Change these to factors
# dats$MARRIED<- as.factor(as.character(dats$MARRIED))
# dats$APPROVE<- as.factor(as.character(dats$APPROVE))
# dats$MALE<- as.factor(as.character(dats$MALE))
# dats$GDLIN<-  as.factor(dats$GDLIN)
# dats$BLACK<-  as.factor(dats$BLACK)
# dats$HISPAN<-  as.factor(dats$HISPAN)

#Removed 20 ob
#Changed it to integer

dats$MARRIED<- as.integer(as.character(dats$MARRIED))
dats$MALE<- as.integer(as.character(dats$MALE))

#dats$GDLIN<-  as.factor(dats$GDLIN)
#attach the file for use
attach(dats)

#Review selected data
glimpse(dats)
kable(summary(dats))
```

##Review GGpair Matrix

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

ggpairs(dats,lower = list(continuous="smooth"))
```

##Review Stats of each variables 

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
##The describe() function which is part of the Hmisc package displays the following additional statistics:
#Number of rows,#Standard deviation,#Trimmed mean,#Mean absolute deviation,#Skewness,#Kurtosis,#Standard error
describe(dats)
```

##Review Variance for anything that may jump out

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
##The stat.desc() function which is part of the pastecs package displays the following additional statistics:
#Variance,#Coefficient of variation,#Confidence interval for mean
kable(stat.desc(dats))
```

##Histogtam of the Variables 

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#Lot more married obs 
plot_ly(x = dats$MARRIED,
             type = "histogram")
#Small amount of obs that desn't meet credit guideline
plot_ly(x = dats$GDLIN,
             type = "histogram")
#Right Skewed with 36 in the middle 
plot_ly(x = dats$OBRAT,
             type = "histogram")
#Small amount of Black obs
plot_ly(x = dats$BLACK,
             type = "histogram")
#Small amount of Hispanic obs
plot_ly(x = dats$HISPAN,
             type = "histogram")
#Mostly Males
plot_ly(x = dats$MALE,
             type = "histogram")
#Mostly Approved
plot_ly(x = dats$APPROVE,
             type = "histogram")
#LOANPRC = loan amount/purchase price
#.80 is 20% of the data, .9 is 10% of the data
plot_ly(x = dats$LOANPRC,
             type = "histogram") 
```

##Quantile-Quantile (QQ) Plots


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# after 2 its not normal 
ggplot(dats, aes(sample = LOANPRC)) + 
  geom_qq() +
  ggtitle("Q-Q Plot for LOANPRC")

#looks normally distrubuted
ggplot(dats, aes(sample = OBRAT)) + 
  geom_qq() +
  ggtitle("Q-Q Plot for OBRAT")
```

##Visulize the relationship & Review Logit relationships

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# print the two graph side-by-side - Approve
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,dats$APPROVE,logi.mod=1, mainlabel="LoanPRC&Approve",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,dats$APPROVE,logi.mod=1, mainlabel="OBRAT&Approve",boxp=FALSE,type="hist",col="gray")
# print the two graph side-by-side - BLACK
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,dats$BLACK,logi.mod=1, mainlabel="LoanPRC&BLACK",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,dats$BLACK,logi.mod=1, mainlabel="OBRAT&BLACK",boxp=FALSE,type="hist",col="gray")
# print the two graph side-by-side - Hispanic
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,dats$HISPAN,logi.mod=1, mainlabel="LoanPRC&Hispanic",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,dats$HISPAN,logi.mod=1, mainlabel="OBRAT&Hispanic",boxp=FALSE,type="hist",col="gray")
```

###Black relationshps and Logit interactions

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# Loan to Purchase Ratio, Other obligations and Approval side-by-side - BLACK
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,(dats$APPROVE*dats$BLACK),logi.mod=1, mainlabel="LoanPRC&Approve",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,(dats$APPROVE*dats$BLACK),logi.mod=1, mainlabel="OBRAT&Approve",boxp=FALSE,type="hist",col="gray")

# Loan to Purchase Ratio, Other obligations and Approval and meets guidline side-by-side - BLACK
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,(dats$APPROVE*dats$BLACK*dats$GDLIN),logi.mod=1, mainlabel="LoanPRC&Approve",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,(dats$APPROVE*dats$BLACK*dats$GDLIN),logi.mod=1, mainlabel="OBRAT&Approve",boxp=FALSE,type="hist",col="gray")

# Loan to Purchase Ratio, Other obligations and Approval and meets guidline and male side-by-side - BLACK
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,(dats$APPROVE*dats$BLACK*dats$GDLIN*dats$MALE ),logi.mod=1, mainlabel="LoanPRC&Approve",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,(dats$APPROVE*dats$BLACK*dats$GDLIN*dats$MALE),logi.mod=1, mainlabel="OBRAT&Approve",boxp=FALSE,type="hist",col="gray")

# Loan to Purchase Ratio, Other obligations and Approval and meets guidline and male and marriedside-by-side - BLACK
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,(dats$APPROVE*dats$BLACK*dats$GDLIN*dats$MALE*dats$MARRIED),logi.mod=1, mainlabel="LoanPRC&Approve&",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,(dats$APPROVE*dats$BLACK*dats$GDLIN*dats$MALE*dats$MARRIED),logi.mod=1, mainlabel="OBRAT&Approve",boxp=FALSE,type="hist",col="gray")

```

###Hispanic relationshps and Logit interactions

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# Loan to Purchase Ratio, Other obligations and Approval side-by-side - Hispanic
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,(dats$APPROVE*dats$HISPAN),logi.mod=1, mainlabel="LoanPRC&Approve",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,(dats$APPROVE*dats$HISPAN),logi.mod=1, mainlabel="OBRAT&Approve",boxp=FALSE,type="hist",col="gray")

# Loan to Purchase Ratio, Other obligations and Approval and meets guidline side-by-side - Hispanic
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,(dats$APPROVE*dats$HISPAN*dats$GDLIN),logi.mod=1, mainlabel="LoanPRC&Approve",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,(dats$APPROVE*dats$HISPAN*dats$GDLIN),logi.mod=1, mainlabel="OBRAT&Approve",boxp=FALSE,type="hist",col="gray")

# Loan to Purchase Ratio, Other obligations and Approval and meets guidline and male side-by-side - Hispanic
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,(dats$APPROVE*dats$HISPAN*dats$GDLIN*dats$MALE ),logi.mod=1, mainlabel="LoanPRC&Approve",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,(dats$APPROVE*dats$HISPAN*dats$GDLIN*dats$MALE),logi.mod=1, mainlabel="OBRAT&Approve",boxp=FALSE,type="hist",col="gray")

# Loan to Purchase Ratio, Other obligations and Approval and meets guidline and male and marriedside-by-side - Hispanic
par(mfrow=c(1,2))
logi.hist.plot(dats$LOANPRC,(dats$APPROVE*dats$HISPAN*dats$GDLIN*dats$MALE*dats$MARRIED),logi.mod=1, mainlabel="LoanPRC&Approve&",boxp=FALSE,type="hist",col="gray")
logi.hist.plot(dats$OBRAT,(dats$APPROVE*dats$HISPAN*dats$GDLIN*dats$MALE*dats$MARRIED),logi.mod=1, mainlabel="OBRAT&Approve",boxp=FALSE,type="hist",col="gray")
```

# EDA of different ethnicity

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# filter out for only Black data
datb <-
  dats %>%
  filter(BLACK == 1) 
# filter out for only Hispanic data
dath <-
  dats %>%
  filter(HISPAN == 1)
# filter out for only Hispanic data
datw <-
  dats %>%
  filter(BLACK == 0) %>%
  filter(HISPAN == 0)
```

## Descriptive Stats Table of different ethnicity
+ Black and Hispanic makes up relatively small proportion of the data set
+ Black and Hispanic has the lower approval rate compared to White

```{r results='asis', echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

stargazer(datw, type = "html", title="Descriptive statistics - White Only", digits=2)
stargazer(datb, type = "html", title="Descriptive statistics - Black Only", digits=2)
stargazer(dath, type = "html", title="Descriptive statistics - Hispanic Only", digits=2)

```
#logit models comparison
Black only model has only one (meets guidelines) significant variable

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

## fit ordered logit model
glm1 <- glm(APPROVE ~ MALE + GDLIN + LOANPRC + OBRAT + MARRIED + BLACK + HISPAN, data = dats, family = binomial(link="logit"))
glm2 <- glm(APPROVE ~ GDLIN + LOANPRC + OBRAT + MARRIED + BLACK + HISPAN, data = dats, family = binomial(link="logit"))
glm3 <- glm(APPROVE ~ GDLIN + LOANPRC + OBRAT + BLACK + HISPAN, data = dats, family = binomial(link="logit"))
glmw <- glm(APPROVE ~ GDLIN + LOANPRC + OBRAT + MARRIED + BLACK + HISPAN, data = datw, family = binomial(link="logit"))
glmb <- glm(APPROVE ~ GDLIN + LOANPRC + OBRAT + MARRIED + BLACK + HISPAN, data = datb, family = binomial(link="logit"))
glmh <- glm(APPROVE ~ GDLIN + LOANPRC + OBRAT + MARRIED + BLACK + HISPAN, data = dath, family = binomial(link="logit"))

## view a summary of the Full model
summary(glm1)
#Removed Male since there is no significance 
summary(glm2)
# Removed Married since there is small significance
summary(glm3)
# White only data
summary(glmw)
# Black Only data
summary(glmb)
# Hispanic only data
summary(glmh)
#Compare the full data set
anova(glm1, glm2, glm3)
```
```{r results='asis', echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

#Compare all the modles 
stargazer(glm1, glm2, glm3, glmw, glmb, glmh, type = "html", 
          column.labels = c("full Data", "white only", 
                            "black only", "hispanic only"), column.separate = c(3, 1,1,1))
          
```


## Odds Ratio for all 6 models


```{r results='asis', echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
OR <- function(x) exp(x)
stargazer(glm1, glm2, glm3, type="html", apply.coef = OR, 
          title = "Odds Ratio with full data set")

stargazer(glmw, glmb, glmh, type="html",
          title = "Odds Ratio with different ethnicity",
          apply.coef = OR, column.labels = c("white only", 
                                             "black only", "hispanic only"))

stargazer(glm1, glm2, glm3, glmw, glmb, glmh, type="html",
          title = "Logistic Regression with t stas for significance",
          report = "vct*", column.labels = c("full Data", "white only",
                                             "black only", "hispanic only"), 
          column.separate = c(3, 1,1,1))
```

```{r results='asis', echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
stargazer(
exp(cbind(OR = coef(glm1), confint(glm1))),
exp(cbind(OR = coef(glm2), confint(glm2))),
exp(cbind(OR = coef(glm3), confint(glm3))),
exp(cbind(OR = coef(glmw), confint(glmw))),
exp(cbind(OR = coef(glmb), confint(glmb))),
exp(cbind(OR = coef(glmh), confint(glmh))),
type="html")
```

## Likelihood Ratio Test

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
lrtest(glm2, glm1)
LR.s.1 <- glm2$deviance - glm1$deviance
LR.s.2 <- -2*logLik(glm2)[1] - (-2*logLik(glm1)[1])
pchisq(LR.s.1, 1, lower.tail = FALSE)
pchisq(LR.s.1, 2, lower.tail = FALSE)
lrm(glm2)
lrm(glm1)
```

## Visulizing Model Fit Analysis 
- Model with male and married committed seems to be the best fit. 

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#FUll Model 
par(mfrow=c(2,2)) # init 4 charts in 1 panel
plot(glm1)
#Ommited Gender
par(mfrow=c(2,2)) # init 4 charts in 1 panel
plot(glm2)
#Ommited Gender and Married
par(mfrow=c(2,2)) # init 4 charts in 1 panel
plot(glm3)
```
 
## Correctly classified observations

```{r results='asis', echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
###Correctly classified observations
#all the same
stargazer(mean((glm1$fitted.values>=0.5)==dats$APPROVE),
mean((glm2$fitted.values>=0.5)==dats$APPROVE),
mean((glm3$fitted.values>=0.5)==dats$APPROVE), type = "text")
```

##Confusion matrix
Same outcome

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

###Confusion matrix count

#FUll Model
RP=sum((glm1$fitted.values>=0.5)==dats$APPROVE & dats$APPROVE==1)
FP=sum((glm1$fitted.values>=0.5)!=dats$APPROVE & dats$APPROVE==0)
RN=sum((glm1$fitted.values>=0.5)==dats$APPROVE & dats$APPROVE==0)
FN=sum((glm1$fitted.values>=0.5)!=dats$APPROVE & dats$APPROVE==1)
confMat1<-matrix(c(RP,FP,FN,RN),ncol = 2)
colnames(confMat1)<-c("Pred Approved","Pred not Approved")
rownames(confMat1)<-c("Real Approved","Real not Approved")
#Ommited Gender
RP=sum((glm2$fitted.values>=0.5)==dats$APPROVE & dats$APPROVE==1)
FP=sum((glm2$fitted.values>=0.5)!=dats$APPROVE & dats$APPROVE==0)
RN=sum((glm2$fitted.values>=0.5)==dats$APPROVE & dats$APPROVE==0)
FN=sum((glm2$fitted.values>=0.5)!=dats$APPROVE & dats$APPROVE==1)
confMat2<-matrix(c(RP,FP,FN,RN),ncol = 2)
colnames(confMat2)<-c("Pred Approved","Pred not Approved")
rownames(confMat2)<-c("Real Approved","Real not Approved")
#Ommited Gender and Married
RP=sum((glm3$fitted.values>=0.5)==dats$APPROVE & dats$APPROVE==1)
FP=sum((glm3$fitted.values>=0.5)!=dats$APPROVE & dats$APPROVE==0)
RN=sum((glm3$fitted.values>=0.5)==dats$APPROVE & dats$APPROVE==0)
FN=sum((glm3$fitted.values>=0.5)!=dats$APPROVE & dats$APPROVE==1)
confMat3<-matrix(c(RP,FP,FN,RN),ncol = 2)
colnames(confMat3)<-c("Pred Approved","Pred not Approved")
rownames(confMat3)<-c("Real Approved","Real not Approved")
###Confusion matrix count for the 3 Models
kable(
  rbind(
    confMat1,
    confMat2,
    confMat3))

###Confusion matrix proportion

#FUll Model
RPR=RP/sum(dats$APPROVE==1)*100
FPR=FP/sum(dats$APPROVE==0)*100
RNR=RN/sum(dats$APPROVE==0)*100
FNR=FN/sum(dats$APPROVE==1)*100
confMat1<-matrix(c(RPR,FPR,FNR,RNR),ncol = 2)
colnames(confMat1)<-c("Pred Approved","Pred not Approved")
rownames(confMat1)<-c("Real Approved","Real not Approved")
#Ommited Gender
RPR=RP/sum(dats$APPROVE==1)*100
FPR=FP/sum(dats$APPROVE==0)*100
RNR=RN/sum(dats$APPROVE==0)*100
FNR=FN/sum(dats$APPROVE==1)*100
confMat2<-matrix(c(RPR,FPR,FNR,RNR),ncol = 2)
colnames(confMat2)<-c("Pred Approved","Pred not Approved")
rownames(confMat2)<-c("Real Approved","Real not Approved")
#Ommited Gender and Married
RPR=RP/sum(dats$APPROVE==1)*100
FPR=FP/sum(dats$APPROVE==0)*100
RNR=RN/sum(dats$APPROVE==0)*100
FNR=FN/sum(dats$APPROVE==1)*100
confMat3<-matrix(c(RPR,FPR,FNR,RNR),ncol = 2)
colnames(confMat3)<-c("Pred Approved","Pred not Approved")
rownames(confMat3)<-c("Real Approved","Real not Approved")

###Confusion matrix proportion
kable(
  rbind(
    confMat1,
    confMat2,
    confMat3))

```

## Prototypical Analysis
Both Probit and Logit prediction shows that Hispanics and Blacks are less likely to get approval to loans when compared to whites
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

#Estimate Logit Model
LogitModel = glm(APPROVE ~ OBRAT + BLACK + HISPAN, data = dats, 
                 family = "binomial")
summary(LogitModel)

#Generate Odds Ratios
exp(coef(LogitModel))

#Define prototypical loan applicants (you will need more than 3)
prototype1 <- data.frame(OBRAT=mean(dats$OBRAT),BLACK = 1, HISPAN = 0)
prototype2 <- data.frame(OBRAT=mean(dats$OBRAT),BLACK = 0, HISPAN = 1)
prototype3 <- data.frame(OBRAT=mean(dats$OBRAT),BLACK = 0, HISPAN = 0)

#Predict probabilities for prototypical individuals
prototype1$predictedprob <- predict (LogitModel, newdata = prototype1, type ="response")
prototype2$predictedprob <- predict (LogitModel, newdata = prototype2, type ="response")
prototype3$predictedprob <- predict (LogitModel, newdata = prototype3, type ="response")
#logit Probability
kable(rbind( prototype1,
       prototype2,
       prototype3))


#Estimate Probit Model
ProbitModel = glm(APPROVE ~ OBRAT + BLACK + HISPAN, data = dats, 
                  family = "binomial" (link = "probit"))
summary(ProbitModel)

#Predict probabilities for prototypical individuals
prototype1$predictedprob <- predict (ProbitModel, newdata = prototype1, type ="response")
prototype2$predictedprob <- predict (ProbitModel, newdata = prototype2, type ="response")
prototype3$predictedprob <- predict (ProbitModel, newdata = prototype3, type ="response")
# Probit Probability
kable(rbind(prototype1,
      prototype2,
      prototype3))

#Impose appropriate sample selection criteria here
MLDsubsample <- subset(dats, LOANPRC >= 50 & MALE != "." & MARRIED != "." & (GDLIN == 1 | GDLIN == 0))

 
#Estimate Logit Model (I am not necessarily recommending this model -- this is purely illustrative)
LogitModel <- glm(APPROVE ~ OBRAT + GDLIN + LOANPRC + MALE + MARRIED + BLACK + HISPAN, data = MLDsubsample, 
                 family = "binomial")
summary(LogitModel)

#Generate Log-Likelihood
logLik(LogitModel)


```

#Probit Model Analysis
Stick with Logit Model since its easier to translate

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

# this gives main effects AND interactions
ad_aov <- aov(APPROVE ~ OBRAT * BLACK + HISPAN * OBRAT + LOANPRC * BLACK + LOANPRC * HISPAN, data = dats, family = "binomial" (link = "probit"))
# this would give ONLY main effects
ad_aov2 <- aov(APPROVE ~ OBRAT + BLACK + HISPAN + MARRIED + LOANPRC + GDLIN, data = dats, family = "binomial" (link = "probit"))
summary(ad_aov)
summary(ad_aov2)
tidy_ad_aov <- broom::tidy(ad_aov)
kable(tidy_ad_aov)

```


